Inequalities by ethnicity in the CHD pathway

### 2024-04-23 findings.qmd
update

---
title: "Visualising CHD Pathway Inequalities by ethnicity - DRAFT"
lang: en-GB
author: "Jacqueline Grout"
date: last-modified
date-format: "YYYY-MM-DD"
title-block-banner: "#f9bf07"
title-block-banner-color: "#333739"
format:
  html:
    self-contained: true
    grid:
      sidebar-width: 200px
      body-width: 950px
      margin-width: 150px
      gutter-width: 1.5rem
    embed-resources: true
    smooth-scroll: true
    theme: cosmo
    fontcolor: black
    toc: true
    toc-location: left
    toc-title: Contents
    toc-depth: 3
editor: visual
execute:
  echo: false
  message: false
  warning: false
  freeze: auto
editor_options: 
  chunk_output_type: console

css: styles.css
---

@@

library(targets)
library(gt)
library(tidyverse)
library(grid)
library(gridExtra)

# Introduction

In 2022, the Strategy Unit worked with the British Heart Foundation (BHF) to explore ways of visualising socio-economic inequalities as they emerge through the coronary heart disease (CHD) pathway. We made these visualisations available so that health and care staff can understand where on the pathway socio-economic inequalities emerge and at which points they are moderated or exacerbated. The report[^1] and web-based tool[^2] are available via the BHF and Strategy Unit websites.

[^1]: <https://www.strategyunitwm.nhs.uk/publications/socio-economic-inequalities-coronary-heart-disease>

[^2]: <https://www.bhf.org.uk/icb-tool>

Whilst this initial analysis was focused on inequalities between socio-economic groups, BHF asked the Strategy Unit if we might explore the feasibility of assessing inequalities across other dimensions of inequality. This report explores inequalities by ethnicity along the coronary heart disease (CHD) pathway.

The objectives of the report are to:

1.  Set out the methods by which the Strategy Unit have sought to represent CHD pathway inequalities by ethnicity.

2.  Quantify and illustrate the inequalities by ethnicity over the disease progression and treatment pathway for CHD.

The analysis has been conducted by the Strategy Unit on behalf of the British Heart Foundation.

# Inequities in healthcare

The term ‘inequities’ is used to describe unjustifiable differences in rates of access between subgroups. An equity analysis must control for levels of need within each population subgroup. Having done this, an equitable distribution of services is one where rates of access to a service or population follow the distribution of need, such that a patient with a given level of need in one subgroup has the same chance of accessing a service as their counterparts with a similar level of need in other subgroups. This is the standard that the NHS seeks to achieve. Assessing equity is challenging. Further detail about inequalities and inequities in healthcare can be found in a previous Strategy Unit report[^3]

[^3]: <https://www.strategyunitwm.nhs.uk/publications/socio-economic-inequalities-access-planned-hospital-care-causes-and-consequences>

In our previous work, visualising socio-economic inequalities, our units of analysis were GP practices. The metrics of interest, risk factors, primary prevention interventions, secondary care procedures, outcomes etc, are readily available at GP practice level and there are established methods of assigning GP practices to deciles of deprivation. Having calculated a metric value for each decile of GP practices, we used the relative index of inequality (RII) to estimate the scale and direction of inequality. The RII can only be used when the dimension of inequality can be expressed as a set of order groups.

Assessing inequalities across ethnic groups is more challenging, since unlike deprivation, ethnicity cannot be expresseed as a set of ordered groups. The distribution of patients across ethnic groups is unique to each practice. There is no meaningful way to numerically aggregate these distributions into a single variable (e.g. % BME) without significant loss of information. An alternative method of grouping practices according to the distribution of its patients over ethnic group was required. We have used k-medoids clustering to assign practices to one of a small number of groups such that practices within a group have similar distributions of patients over ethnic groups. K-medoids is an established, and commonly used unsupervised machine learning technique. This clustering approach required data on the ethnicity of a practices registered population. We imputed it based on the Lower Super Output Area (LSOA) of residence of GP registrants and the ethnicity distribution of patients living in these LSOAs using the Census 2021 data.

Ethnicity, unlike deprivation, is not an ordered variable. The approach in our previous work used the relative index of inequality to measure the degree of inequality. This measure relies on the ordered quality of the socio-economic deprivation variable. To handle the categorical nature of the ethnicity variable we have used the relative index of disparity to indicate the extent to which the rate of an activity or event varies across groups. The index estimates the proportion of events (e.g., admissions) that would need to be redistributed between clusters in order that event rates follow levels of need. Further detailed explanations of the methods used in the analysis can be found in the appendix.

# CHD Metrics

This report quantifies and illustrates levels of inequity across 35 metrics at various points along the continuum of coronary heart disease progression and over a typical treatment pathway. They are shown in the table below grouped by domain (risk factors, risk factor identification, primary prevention, disease identification, secondary prevention, tertiary prevention, intermediate and full outcomes), which represent the various stages along the pathway. Full definitions and data sources for each metric are included in the appendix.

*Table 1 - coronary heart disease pathway metrics*

```{r}
library(gt)
metrics_table <- tibble(metric_domain = c("Need","Risk factors","Risk factor identification","Risk factor identification","Risk factor identification","Risk factor identification","Risk factor identification","Risk factor identification","Primary prevention","Primary prevention","Primary prevention","Primary prevention","Disease identification","Disease identification","Disease identification","Secondary prevention","Secondary prevention","Secondary prevention","Secondary prevention","Secondary prevention","Secondary prevention","Secondary prevention","Secondary prevention","Tertiary prevention","Tertiary prevention","Tertiary prevention","Tertiary prevention","Tertiary prevention","Tertiary prevention","Intermediate outcome","Intermediate outcome","Intermediate outcome","Full outcomes","Full outcomes","Full outcomes","Full outcomes"),
  name_metric=c("CHD synthetic prevalence estimates",
                                      "Smoking synthetic prevalence estimates",
                                      "Smoking register",
                                      "Obesity register",
                                      "Diabetes register",
                                      "Depression register",
                                      "The percentage of patients aged 45 or over who have a record of blood pressure in the preceding 5 years",
                                      "CVD risk register",
                                      "Percentage of patients aged 18 and over with GP recorded CVD (narrow definition), who are currently treated with lipid lowering therapy",
                                      "Percentage of patients aged 18 and over, with no GP recorded CVD and a GP recorded QRISK score of 10% or more, CKD (G3a to G5), T1 diabetes (aged 40 and over) or T2 diabetes aged 60 and over, who are currently treated with lipid lowering therapy.",
                                      "Smoking cessation support offered",
                                      "Exception reporting for Smoking cessation support offered",

                                      "CHD register",
                                      "CT angiography",
                                      "Electrocardiography",
                                      "Aspirin, anti-platelet or anti-coagulent",
                                      "Exception reporting for Aspirin, anti-platelet or anti-coagulent",
                                      "Flu vaccination",
                                      "Exception reporting for Flu vaccination",
                                      "Percentage of patients aged 65 or over
who received a seasonal influenza vaccination between 1 September 2022 and 31 March 2023",
                                      " Percentage of patients aged 18 to 64
years and in a clinical at-risk group who received
a seasonal influenza vaccination between 1
September 2022 and 31 March 2023",
"Referral to cardiology (First outpatient)",
                                      "Cardiology outpatient DNAs",
"Elective PCI","Elective CABG","Waiting time for elective PCI / CABG","Elective PCI / CABG patients discharged before trimpoint",
"Cardiac rehabilitation - Started", "Cardiac rehabilitation - Completed","BP reading < 140/90","Readmission within 30 days of elective PCI / CABG","Emergency admissions for CHD","Deaths in hospital from CHD","Deaths in hospital from CHD <75","Deaths from CHD","Deaths from CHD <75"
                                      )
                        )

metrics_table |>
    gt()|>
  cols_hide(metric_domain)|>
 # tab_stubhead(label = md("**Domain**"))|>
    tab_header(
    title = md("**Table 1** - *Coronary heart disease pathway metrics*")
  )|>
   tab_row_group(
     id="Need",
    label = md("**Need**"),
    rows = 1
  ) |>
  tab_row_group(
    id="Risk factors",
    label = md("**Risk factors**"),
    rows = 2
  ) |>
  tab_row_group(
    id="Risk factor identification",
    label = md("**Risk factor identification**"),
    rows = 3:7
  )|>
  tab_row_group(
    id="Primary prevention",
    label = md("**Primary prevention**"),
    rows = 8:12
  )|>
  tab_row_group(
    id="Disease identification",
    label = md("**Disease identification**"),
    rows = 13:15
  )|>
  tab_row_group(
    id="Secondary prevention",
    label = md("**Secondary prevention**"),
    rows = 16:23
  )|>
    tab_row_group(
    id="Tertiary prevention",
    label = md("**Tertiary prevention**"),
    rows = 24:29
  )|>
    tab_row_group(
    id="Intermediate outcome",
    label = md("**Intermediate outcome**"),
    rows = 30:32
  )|>
    tab_row_group(
    id="Full outcomes",
    label = md("**Full outcomes**"),
    rows = 33:36
  )|>
  row_group_order(groups=c("Need","Risk factors","Risk factor identification","Primary prevention","Disease identification","Secondary prevention","Tertiary prevention","Intermediate outcome","Full outcomes"))|>
  cols_label(name_metric="")

::: {.callout-note icon="false"}
## Cardiac rehabilitation metrics

Cardiac rehabilitation metrics will be added once they become available from NACR (expected before 31/3/24)
:::

# Clustering

In this analysis we used the k-medoids clustering method to group practices. Each practice was assigned to one of a small number of groups such that practices within the group have similar distributions of patients over ethnic groups. In order to obtain data on the ethnicity of a practice's registered population we imputed it based on the Lower Super Output Area (LSOA) of residence of GP registrants and the ethnicity distribution of patients living in those LSOAs using the Census 2021 data. Further detailed explanations of the method can be found in the appendix.

The clustering analysis resulted in 5 groups (clusters) of practices which are illustrated in the charts and map below.

## Cluster descriptions

The median percentage of each ethnic group for each cluster was calculated and is presented in these charts to describe their respective diversity.

### Cluster 1 - Least diverse

@@

::: panel-tabset
## Five ethnic groups

@@

## Fourteen ethnic groups

@@

The interactive map below presents each GP practice coloured according to their assigned cluster. Cluster 1 (red) is the least diverse and cluster 5 (orange) is the most diverse.

# Rate of activity

The charts below show the activity rates by cluster for each metric along the CHD pathway. The horizontal line is the overall rate. The 95% confidence intervals are shown at the top of each bar. These activity rates take account of need by using CHD prevalence (or list size for risk factor) as the denominator of the calculation. Further detail on the methodology can be found in the appendix.

In some cases these upper and lower limits are so small as to not be visible (e.g. Obesity register), whereas for other metrics (e.g. Readmission within 30 days of a PCI or CABG) they are much wider. Metrics with larger volumes of activity data give rise to a greater confidence in the calculated rate (e.g. limits are closer together) when compared to metrics with smaller volumes of activity data where there is less confidence in the calculated rate (e.g. limits are further apart).

In calculating these rates for each cluster within each metric and then comparing the rates by cluster to the global rate for the metric it is possible to identify how activity rates vary across the 5 clusters. For there to be equity between the clusters the rates would need to be the same. The charts below present the metrics grouped according to pathway domain.

@@

::: callout-tip
## Key Findings

-   Activity rates vary very little for the risk factor metric % patients 45+ with a record of BP \< 5 years

-   Rates of patients on the diabetes register are highest in the most diverse cluster

-   Rates of patients on the depression register are highest in the least diverse cluster and lowest in the most diverse cluster
:::

#### Risk factors

@@

#### Risk factor identification

@@

::: callout-tip
## Key Findings

-   The activity rate for CVD patients treated with LLT is highest for those patients whose practice is in the least diverse cluster and lowest for patients whose practice is in cluster 4.

-   A similar pattern applies for at risk patients treated with LLT, activity rates being highest for those in the least diverse cluster.
:::

@@

::: callout-tip
## Key Findings

-   Activity rates vary considerably across the clusters in the disease identification metrics, with differing patterns for each metric

-   Patients from clusters 4 and 5 (most diverse) are less likely to be recorded on the CHD register

-   Clusters 5 and 4 have lower rates of CT angiography and electrocardiography
:::

@@

::: callout-tip
## Key Findings

-   There is little variation between clusters 1 to 4 in referral rates to outpatient cardiology, although the referrals are lower for the most diverse cluster (cluster 5)

-   Flu vaccination rates for 65+ patients are highest among patients whose practice is in the least diverse cluster. the two most diverse clusters having the lowest rates

-   Amongst patients under 65 who are at risk, flu vaccination rates are less varied between the clusters than for 65+ patients. Cluster 4 has the lowest rate of vaccination

-   Cardiology outpatient DNA rates are higher the more diverse the cluster
:::

@@

::: callout-tip
## Key Findings

-   All 4 tertiary prevention metrics follow the same pattern of activity rates with the highest rates being in the least diverse cluster (cluster 1), followed by clusters 2 and then cluster 5 (the most diverse) and cluster 3. Cluster 4 has the lowest activity rates.
:::

@@

::: callout-tip
## Key Findings

-   The least diverse cluster has the highest rates of BP readings \< 140/90.

-   Emergency admissions for CHD and readmissions within 30 days of a PCI or CABG are highest for the least diverse cluster and lowest for cluster 4.
:::

@@

::: callout-tip
## Key Findings

...to be determined - awaiting death data check
:::

@@

# Relative Index of Disparity

The relative index of deprivation for the metric is expressed here as a %, such that the % indicates the amount of activity that would need to be redistributed between clusters in order that event rates follow levels of need. The detail of the methodology used for its calculation can be found in the appendix.

The chart below presents the metrics along the CHD pathway from risk factor and identification through to outcome measures of death from CHD. For each metric the index of disparity is shown as a point estimate (yellow dot) with the upper and lower confidence intervals in grey.

::: callout-tip
## Key Findings

-   The greatest disparity is flu vaccinations for patients aged 65+ where the relative index of disparity is 16.70% (95% ci - 16.64%, 16.70%), indicating that 16.7% of activity needs to be redistributed so that rates reflect need.
:::

::: {.callout-tip icon="false"}
## Confidence Intervals

Confidence intervals are shown in grey. There is 95% confidence that the index of disparity is within this range. The confidence interval estimation methods are detailed in the appendix.
:::

## Routes to equity

A previous Strategy Unit report "Strategies to reduce inequalities in access to planned hospital procedures"[^4] highlights in chapter 1, three of the many routes from inequity to equity: levelling-up, levelling-down and zero-sum redistribution. Using the zero-sum redistribution route as an example, along with the metric flu vaccinations for patients aged 65+, this illustration shows how equity could be achieved by increasing and decreasing the number of vaccinations delivered in each cluster to give the same activity rate per cluster.

[^4]: <https://www.midlandsdecisionsupport.nhs.uk/knowledge-library/strategies-to-reduce-inequalities-in-access-to-planned-hospital-procedures/>

|  Cluster  | Number of vaccinations | CHD synthetic prevalence | Rate (No. of vaccinations / CHD synthetic prevalence) | Global rate |  Change in number of vaccinations to match global rate  |
|:----------:|-----------:|-----------:|:----------:|:----------:|:----------:|
|           |                    *n* |                      *d* |                        *r=n/d*                        | *gr=𝚺n/𝚺d*  |                      *(r-gr) x d*                       |
|     1     |              3,356,002 |                1,026,826 |                         3.27                          |    2.33     |                        -961,239                         |
|     2     |              2,638,698 |                  924,088 |                         2.86                          |    2.33     |                        -483,540                         |
|     3     |              1,862,430 |                  958,825 |                         1.94                          |    2.33     |                        +373,741                         |
|     4     |                476,616 |                  478,728 |                         1.00                          |    2.33     |                        +639,873                         |
|     5     |                332,728 |                  327,542 |                         1.02                          |    2.33     |                        +431,166                         |
| **Total** |          **8,666,474** |            **3,716,009** |                       **2.33**                        |             |                          **0**                          |

: In reality, in attempting to achieve equity for this metric, it is unlikely that the offer of vaccination would be removed from patients whose practice is in clusters 1 and 2. More likely would be an approach to focus public health and vaccination roll-out campaigns on practices in clusters 3, 4 and 5.

Each of the metrics along the CHD pathway with the higher relative indicies of disparity would almost certainly require a variety of different routes from inequity to equity given their differing nature. The more effective interventions on the pathway might be levelled-up, whilst those of more limited value might be levelled-down.

# Regional Analysis

## Clusters

The charts below show the distribution of the patients in each NHS region according to the cluster of their GP practice.

::: panel-tabset
## London

```{r}
#1
region_cluster_charts <- tar_read(region_cluster_charts)
region_cluster_charts[2]
```

## South West

```{r}
#2
region_cluster_charts[7]
```

## South East

```{r}
#3
region_cluster_charts[6]
```

## Midlands

```{r}
#4
region_cluster_charts[3]
```

## East of England

```{r}
#5
region_cluster_charts[1]
```

## North West

```{r}
#6
region_cluster_charts[5]
```

## North East and Yorkshire

```{r}
#7
region_cluster_charts[4]
```
:::

::: callout-tip
## Key Findings

-   The greatest disparity
:::

::: panel-tabset
## London

```{r}
regional_charts <- tar_read(regional_charts)
regional_charts[1]
```

## South West

```{r}
regional_charts[2]
```

## South East

```{r}
regional_charts[3]
```

## Midlands

```{r}
regional_charts[4]
```

## East of England

```{r}
regional_charts[5]
```

## North West

```{r}
regional_charts[6]
```

## North East and Yorkshire

```{r}
regional_charts[7]
```
:::

The appendix contains equivalent charts to those above, showing the Index of Disparity for each CHD pathway metric for each of the 42 ICBs.

# Conclusions

.......

# Appendix

## Definitions and data sources of pathway metrics

The table below sets out the sources of the various metrics as well as the time-period to which they relate, and details of the selection criteria used.

| Domain                     | Metric                                 | Data Source | Year    | Definition and selection criteria/codes                                                 |
|---------------|---------------|---------------|---------------|---------------|
| Need                       | CHD synthetic prevalence estimates     | PHE         | 2015    | Provided directly by OHID (formerly indicator 92847 in FingertipsR)                     |
| Risk factors               | Smoking synthetic prevalence estimates | GP survey   | 2023    |                                                                                         |
| Risk factor identification | Smoking register                       | NHSD QOF    | 2022/23 | [Indicator 91280 extracted fromFingertipsR](https://fingertips.phe.org.uk/search/91280) |
|                            | Obesity register                       | NHSD QOF    |         | [Indicator 93088 extracted fromFingertipsR](https://fingertips.phe.org.uk/search/93088) |
|                            | Diabetes register                      | NHSD QOF    |         |                                                                                         |
|                            | Depression register                    | NHSD QOF    |         |                                                                                         |
| Primary prevention         |                                        |             |         |                                                                                         |
|                            |                                        |             |         |                                                                                         |
|                            |                                        |             |         |                                                                                         |
|                            |                                        |             |         |                                                                                         |
|                            |                                        |             |         |                                                                                         |
|                            |                                        |             |         |                                                                                         |
|                            |                                        |             |         |                                                                                         |
|                            |                                        |             |         |                                                                                         |
|                            |                                        |             |         |                                                                                         |
|                            |                                        |             |         |                                                                                         |
|                            |                                        |             |         |                                                                                         |
|                            |                                        |             |         |                                                                                         |
|                            |                                        |             |         |                                                                                         |
|                            |                                        |             |         |                                                                                         |
|                            |                                        |             |         |                                                                                         |
|                            |                                        |             |         |                                                                                         |
|                            |                                        |             |         |                                                                                         |
|                            |                                        |             |         |                                                                                         |
|                            |                                        |             |         |                                                                                         |
|                            |                                        |             |         |                                                                                         |
|                            |                                        |             |         |                                                                                         |
|                            |                                        |             |         |                                                                                         |
|                            |                                        |             |         |                                                                                         |
|                            |                                        |             |         |                                                                                         |
| Full Outcomes              | Deaths in hospital from CHD            |             |         |                                                                                         |
|                            | Deaths in hospital from CHD \< 75      |             |         |                                                                                         |
|                            | Deaths from CHD                        |             |         |                                                                                         |
|                            | Deaths from CHD \<75                   |             |         |                                                                                         |

## Methods explained

### QOF data

Metrics taken from the QOF data via Fingertips have been extracted by GP practice and the counts (numerator of the measure) from the calculations have been used. Exception reporting metrics have used the Personalised Care Adjustment (PCA) data to obtain the number of patients by GP practice for whom a PCA has been recorded for the relevant metric. Possible reasons for a PCA include; Newly diagnosed/registered, Intervention is clinically unsuitable, patient choice, did not respond to offers of care, service not available.

### SUS data

SUS data was extracted using Transact-SQL and using relevant OPCS, ICD10, HRG, and Treatment function codes as detailed in table 2 above.

### Cardiac rehabilitation

This data was supplied by NACR. The data wasn’t available by GP Practice, so an alternative method was followed. Practices were allocated to a cluster and data was then supplied aggregated into clusters. In some cases, the data had to be redacted due to small numbers in the category. For the purposes of this analysis the redacted data item was reassigned an average value. The true definition of eligible for rehabilitation is the HES admission and successful discharge of all patients with CVD e.g., MI, revascularization, or HF. However, to link with the NACR only those records entered into the database can be used so NACR use pseudo eligible as this is as close as the audit can get, at a granular level, to eligible. It is though affected by areas without entry to the database, The data used relates to the 2023 calendar year, which is the most recent complete year, and following changes in the way rehabilitation has been delivered due to the Covid pandemic, most closely reflects the rehabilitation model now in place.

### Readmission within 30 days

The methodology used for this metric follows, as far as possible, that detailed by NHS Digital[^5]. The readmissions counted relate to those who had an elective PCI / CABG in the period 1/3/2022 to 28/2/2023.

[^5]: <https://digital.nhs.uk/data-and-information/publications/statistical/ccg-outcomes-indicator-set/specifications/3.2-emergency-readmissions-within-30-days-of-discharge-from-hospital_1_4>

### ONS Death records

ONS death records contain an encrypted HES ID which was used to link to outpatient, inpatient and A&E HES records, from which the latest GP practice recorded (on a spine traced record) for the patient was then assigned to the death record.

### CHD synthetic prevalence estimates

Public Health England (PHE) CHD synthetic prevalence estimates data was previously available from OHID via Fingertips. This data has now been removed from Fingertips and archived, however OHID supplied it via email for this project following a request. This data is for 7564 practices from 2015, of which 6297 are current practices. In order to assign a prevalence for practices with missing data a nearest neighbours methodology was used to impute the prevalence by taking an average of the prevalence of each practice within a 1.5km radius of the practice without data. Further details, including code with sample data, for this methodology are available in a blog post on the Strategy Unit Data Science website[^6].

[^6]: <https://the-strategy-unit.github.io/data_science/blogs/posts/2024-01-17_nearest_neighbour.html>

### Ethnicity by GP practice

Ethnicity by GP practice was imputed based on the Lower Super Output Area (LSOA) of residence of GP registrants and the ethnicity distribution of patients living in these LSOAs using the Census 2021 data.

#### GP list size by LSOA

Since metric data is mainly from the 2022/23 year, the October 2022 list size data was taken from the NHS Digital Patients Registered at a GP practice dataset[^7] as this is the data closest to the mid point in the 22/23 year. The GP to ICB mapping file was also downloaded and joined to the list size to easily create analysis by ICB and region.

[^7]: <https://digital.nhs.uk/data-and-information/publications/statistical/patients-registered-at-a-gp-practice/october-2022>

#### Census 2021 Ethnicity by LSOA

The 2021 Census Ethnicity data was downloaded from the .gov.uk website.[^8] Summary details of the census ethnicity data is also available and this includes descriptions of the standardised list of ethnic groups and the differences from the 2011 ethnic groups.[^9]

[^8]: <https://www.ethnicity-facts-figures.service.gov.uk/uk-population-by-ethnicity/demographics/age-groups/latest/#download-the-data>

[^9]: <https://www.ethnicity-facts-figures.service.gov.uk/uk-population-by-ethnicity/demographics/age-groups/latest/>

The 2021 Census Ethnicity data by LSOA uses the 2021 LSOAs. The GP list size data uses the 2011 LSOAs. This results in gaps when the datasets are combined. To overcome this the following method was used:

-   An LSOA lookup showing 2011 and 2021 LSOA parent and child relationships was joined to the 2021 Census Ethnicity data

-   Where a 2021 LSOAs has a parent LSOA in 2011 then we applied the average ethnicity % for the child LSOAs to the parent (this is when one 2011 LSOA has been split into multiple 2021 LSOAs).

-   Where the LSOAs in 2021 have a child LSOA in 2011 then we applied the ethnicity % for the 2021 parent to the 2011 children (this is when several LSOAs have been combined into one new LSOA).

Further details about item editing and imputation processes for Census 2021, England and Wales are available on the ONS website. This details techniques used by ONS with the aim of arriving at a fully populated clean database. These techniques include manual imputation and nearest neighbour donor imputation. For the England and Wales Census 2021 the imputation rate for ethnic group was 1.3%.[^10]

[^10]: <https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/methodologies/itemeditingandimputationprocessforcensus2021englandandwales>

### K-Medoids Clustering

K-medoids is an established, and commonly used unsupervised machine learning technique. This clustering approach required data on the ethnicity of a practices registered population which was obtained as per the methodology explained above (Ethnicity by GP practice). The data used in the clustering was the % of patients in each ethnic group per practice, such that for each practice all the ethnic group %'s added to 100%.

The K-medoids clustering was performed using the {cluster} package[^11] in R using partitioning (clustering) of the data into k clusters "around medoids" (PAM), which is a more robust version of K-means. The algorithm is based on the search for k representative objects or medoids among the observations of the dataset. After finding a set of k medoids, k clusters are constructed by assigning each observation to the nearest medoid. The goal is to find k representative objects which minimise the sum of the dissimilarities of the observations to their closest representative object.[^12]

[^11]: Maechler M, Rousseeuw P, Struyf A, Hubert M, Hornik K (2023). *cluster: Cluster Analysis Basics and Extensions*. R package version 2.1.6 — For new features, see the 'NEWS' and the 'Changelog' file in the package source), [https://CRAN.R-project.org/package=cluster](https://cran.r-project.org/package=cluster).

[^12]: <https://cran.r-project.org/web/packages/cluster/cluster.pdf>

Clustering experiments were conducted using a variety of different combinations of ethnic breakdowns, such as 14 ethnic groups and 5 ethnic groups, as well as variations that excluded the % of White British patients or factored in age group. Techniques were also used to attempt to determine the optimal number of clusters such as the elbow method, gap plots, and scree plots. Consideration was given to combining some ethnic groups and in doing so a correlation matrix was produced to identify ethnic groups that commonly featured together. It was also important to seek a number of clusters that could be defined easily, where characteristics did not overlap each other so that resulting clusters were too similar to each other. It was also necessary to ensure that there was a sufficient number of practices in each cluster, too few practices in a cluster could result in issues with small amounts of metric data when calculating activity rates and consequently lead to wide confidence intervals in the index of disparity calculations.

In conclusion, the final version selected, and presented here, used the percentage of the GP list size in each of the 14 census 2021 ethnic groups and five clusters was felt to give a sufficient number of describable clusters for which the index of disparity calculations could be performed. The chart below is an elbow plot generated for the 14 ethnic groups showing an elbow at around 4 or 5 clusters.

```{r}
tar_read(elbow_plot)
```

Table 3 shows the number of GP practices assigned to each of the 5 clusters and the characteristics of the clusters are described in the main body of the report above.

```{r}
final_data_full_cats_percent_5_clusters <- tar_read(final_data_full_cats_percent_5_clusters) |> as_tibble()

final_data_full_cats_percent_5_clusters |>
  group_by(cluster)|>
  count("practice_code")|>
  ungroup() |>
  select(cluster,n)|>
  gt()|>
    tab_header(title = md("**Table 3** - *Number of practices per cluster*"))|>
    cols_label(cluster=md("**Cluster**"),
             n=md("**Number of practices**"))
```

### Index of Disparity calculations

#### Rate of activity

Calculations of the index of disparity begin by calculating an activity rate using a numerator of the activity ( eg number of elective procedures) divided by a denominator, which for metrics other than risk factors has been taken to be CHD synthetic prevalence (as a determinant of need).

Where the metric is in the risk factor domain the denominator used is GP practice list size (with the exception of the percentage of patients aged 45 or over who have a record of blood pressure in the preceding 5 years, where the 45+ list size has been used).

In calculating these rates for each cluster within each metric and then comparing these rates by cluster to the global rate for the metric it is possible to identify how activity rates vary across the 5 clusters

#### Index of disparity

The relative index of disparity indicates the extent to which the rate of an activity or event varies across groups. Having calculated the rates for each cluster within each metric the disparity is calculated by taking the differences for each cluster from the global rate and then determining the amount of activity by which this cluster varies from the global. This absolute difference is then summed for the 5 clusters and divided by twice the sum of the numerators. This produces a relative index of deprivation for the metric, expressed as a %, such that the % indicates the amount of activity that would need to be redistributed between clusters in order that event rates follow levels of need.

Index of disparity = (𝚺 \| r~(1–n)~ - R\| / n) / R\*100

where r = group rate and R = total population rate

A report by Pearcy and Keppel (2002) presents the Index of Disparity as a summary measure of disparity across population groups and advocates its use for groups defined in terms of categories such as ethnicity.[^13]

[^13]: <https://journals.sagepub.com/doi/epdf/10.1093/phr/117.3.273>

### Confidence Intervals

Confidence intervals of 95% are presented in the national charts presented in this report. Confidence intervals have been used, rather than calculations of statistical significance as this presents a clearer picture of the likelihood of the disparity and due to concerns with the suitability in this case of statistical significance.

> "Practitioners need to be aware that statistical significance differs from practical importance in that statistical significance is highly dependent upon sample size. For a large sample, a statistically significant result is likely to be obtained even when the actual magnitude of an effect is small and of little or no practical importance. On the other hand, for a small sample, it is quite likely that insufficient evidence of a statistically significant result will be obtained – even when there is, indeed, an effect of practical importance." (Hahn, G. J., Doganaksoy, N., Meeker, W. Q., 2019)[^14]

[^14]: <https://academic.oup.com/jrssig/article/16/4/20/7038025>

In order to estimate the confidence intervals for the index of disparity, a bootstrap method was adopted, assuming independence, generating 1000 random samples for each metric for which an index of disparity was calculated. The 2.5th and 97.5th percentiles were then taken for each metric to give the upper and lower limits of the 95% confidence interval for each.

An explanation of bootstrap theory and confidence intervals is available in Practical Statistics for Data Scientists (Bruce, P. and Bruce, A., 2017)[^15]

[^15]: <https://www.oreilly.com/library/view/practical-statistics-for/9781491952955/>

## Integrated Care Board (ICB) Analysis

#### Relative Index of Disparity

```{r}
icb_charts <- tar_read(icb_charts)
icb_charts[1]
icb_charts[2]
icb_charts[3]
icb_charts[4]
icb_charts[5]
icb_charts[6]
icb_charts[7]
icb_charts[8]
icb_charts[9]
icb_charts[10]
icb_charts[11]
icb_charts[12]
icb_charts[13]
icb_charts[14]
icb_charts[15]
icb_charts[16]
icb_charts[17]
icb_charts[18]
icb_charts[19]
icb_charts[20]
icb_charts[21]
icb_charts[22]
icb_charts[23]
icb_charts[24]
icb_charts[25]
icb_charts[26]
icb_charts[27]
icb_charts[28]
icb_charts[23]
icb_charts[29]
icb_charts[30]
icb_charts[31]
icb_charts[32]
icb_charts[33]
icb_charts[34]
icb_charts[35]
icb_charts[36]
icb_charts[37]
icb_charts[38]
icb_charts[39]
icb_charts[40]
icb_charts[41]
icb_charts[42]
```

## R code on Github

All the code used in this project is available on the Strategy Unit GitHub[^16].

[^16]: <https://github.com/The-Strategy-Unit/bhf_chd-ethnicity>

>>>>

---
title: "Inequalities by ethnicity in the CHD pathway"
lang: en-GB
author: "Jacqueline Grout"
date: last-modified
date-format: "YYYY-MM-DD"
format:
  html:
    grid:
      sidebar-width: 250px
      body-width: 800px
      margin-width: 250px
      gutter-width: 1.5rem
    embed-resources: true
    smooth-scroll: true
    theme: cosmo
    fontcolor: black
    toc: true
    toc-location: left
    toc-title: Contents
    toc-depth: 3
editor: visual
execute:
  echo: false
  message: false
  warning: false
  freeze: auto
editor_options: 
  chunk_output_type: console
---

@@

library(targets)

library(tidyverse)
```

### Cluster descriptions

#### Cluster 1 - Least diverse

@@

::: panel-tabset
## Five categories

@@

## Fourteen categories

@@

```{r}
tar_read(rel_iod_chart)
```

## Rate

@@

@@

### Risk factor identification

@@

@@

@@

@@

@@

@@

@@
